Follow-up Samples

### 2017-03-24 submission/manuscript_outline_20161024.Rmd
add discussion to method changes

## Discussion
This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differ and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].  Here we describe how the bacterial community changes after treatment and which OTUs may be most important to lesion classification.  Interestingly, many of the most important OTUs for both treatment and lesion classification had taxonomic identification for resident gut microbes.  This suggest that members within the commensal community may be the first that change during CRC pathogenesis.  These changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].

Unlike previous studies on the microbiome and CRC, ours focuses on identifying commonalities within both adenoma and carcinoma groups before and after treatment.  Although there were differences for genera associated with specific bacterium linked with CRC [Figure S2].  These changes were not consistent across lesion.   Instead, most of our results provide support that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria [Figure 2-3 & Table S3].  Additionally, treatment with chemotherapy or radiation changed these bacteria and provided a larger decrease in positive probability then removal of lesion.

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models. There are several reasons why this may have occurred. First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy. Second, is that our Random Forest models were able to gather the same information from other OTUs. Third, is that changes in commensal bacteria are the common differences across adenoma and carcinoma. Finally, it is also possible that all these explanations could have played a role.  Our observations though, would suggest that an individual's resident bacteria have a large role to play in polyp formation and could change in a way that allows predictive models to lower the positive probability of lesion or initial sample before treatment [Figure 2 & 3]. It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after treatment. It is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after treatment [Figure S2].

Many of the common OTUs that we identified taxnomically classified to potential butyrate producers [Table S2].  Another batch of OTUs classified to bacteria that can either degrade polyphenols or are inhibited by them. Both butyrate and polyphenols are thought to be protective against cancer in part by reducing inflammation [@louis_gut_2014].  These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes. One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.

A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) than carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).  These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time. Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1C]. This confounding though would not affect the observations where these individuals were grouped together [Figures 2-4].

Another limitation was that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis. This information would help to strengthen the case for this model keeping several individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion. This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry. Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations. We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

Interestingly, within the treatment model the test data performed better than the training data. This may have occurred because the training AUC determined from 20 repeated 10-fold cross validation removed samples at random and did not consider that they were matched samples. Another potential reason is that the model itself may have been over-fit since the total number of samples was not that large. However, the lesion model did not suffer from these discrepancies. Further independent studies need to be carried out to verify our findings on lesion and treatment changes due to these limitations.

Despite these shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma and carcinoma treatment. Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of lesion or initial sample after treatment provides evidence that there are specific signatures, mostly attributable to commensal organisms, associated with both treatment and lesion. Our data provides evidence that commensal bacteria may be important in the development of polyps and potentially the transition from adenoma to carcinoma.

@@

***Study Design and Patient Sampling:***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after. The bacterial community has been shown to normalize back to a pre-colonscopy community within this time period [@obrien_impact_2013]. Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

>>>>

## Discussion
This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differ and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].  Here we show that the bacterial community changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide an additive change.  Although some of the important OTUs classified to genera from bacteria considered the usual suspects (e.g. *Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not.  The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, SRN, and carcinoma models.  This suggests that members within the commensal community may be the first that change during CRC pathogenesis.  These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].

Unlike previous studies on the microbiome and CRC, ours focuses both on identifying commonalities and differences within adenoma, SRN, and carcinoma groups before and after treatment.  Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3].  The majority of important OTUs taxonmically classified to commensal bacteria [Figure S3].  Although these changes may be subtle, due to the lack of signifcant difference in the bacterial community before and after treatment in adenoma and SRN [Figure 1], they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.

Many of the common OTUs that we identified taxnomically classified to potential butyrate producers (e.g. *Clostridiales*, *Roseburia*,  and *Anaerostipes*)[Table S2].  Other OTUs classified to bacteria that are inhibited by polyphenols (e.g. *Bacteroides*). Both butyrate and polyphenols are thought to be protective against cancer, in part by reducing inflammation [@louis_gut_2014].  These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes. One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016].  High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016].  Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016].  Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression.

A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or SRN versus carcinoma *(uncorrected P-value 0.05)*, with time passed being less for adenoma *( days)* and SRN *( days)* than carcinoma *( days)*.  These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time. Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1A & 1C]. However, there are two reasons that this may not be the case.  First, the SRN group did not have a significant difference and the collection time of their follow up sample was less than the adenoma group.  Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].

Another limitation was that we do not know whether individuals who were still classified as positive by the carcinoma model eventually had a subsequent CRC recurrence. This information would help to strengthen the case for this model keeping numerous individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having carcinoma. This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry. Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations. We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

Despite these shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma, SRN or carcinoma treatment. Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of carcinoma after treatment provides evidence that there are specific signatures, attributable to both inflammtory and resident commensal organisms, associated with treatment. Our data provides evidence that commensal bacteria may be important in the development of polyps and potentially the transition from adenoma to carcinoma.

@@

***Study Design and Patient Sampling:***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after. The bacterial community has been shown to normalize back to a pre-colonscopy community within this time period [@obrien_impact_2013]. Our training cohort consisted of a total of `r length(rownames(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)))))` individuals [Table 1].  Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 2]. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.
